---
title: "Global Energy Consumption Analysis"
subtitle: "Source: Our World in Data"
author: "Mark Hamer"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup

library(tidyverse)
library(scales)
```

```{r}
#| label: load-data

url <- "https://raw.githubusercontent.com/owid/energy-data/master/owid-energy-data.csv"
energy_raw <- read_csv(url)

glimpse(energy_raw)
```

```{r}
#| label: data-structure

# How many rows and columns?
cat("Rows:", nrow(energy_raw), "\n")
cat("Columns:", ncol(energy_raw), "\n")

# Column names
names(energy_raw)

# Representative sample of raw wide structure
energy_raw |>
  filter(country %in% c("World", "Saudi Arabia")) |>
  filter(year == 2022) |>
  select(country, year, oil_electricity, gas_electricity, 
         solar_electricity, wind_electricity) |>
  knitr::kable(caption = "Sample of raw wide-format data before tidying")
```
## Introduction

Global energy demand is rising — driven by population growth, industrialization,
and the rapid expansion of AI data centers. At the center of this story sits the
**Gulf Cooperation Council (GCC)**: Saudi Arabia, the UAE, Qatar, Kuwait, Bahrain,
and Oman.

These nations face a unique paradox. They occupy one of the **hottest regions on
earth**, creating enormous domestic energy demand for cooling and desalination —
yet they also sit atop the world's largest fossil fuel reserves, making their
economies almost entirely dependent on energy exports.

That dependence makes regional stability an existential economic priority. The
**Strait of Hormuz** — controlled along its northern shore by Iran — is the
world's most critical energy chokepoint. Roughly **20% of global oil supply**
passes through it daily. Any military escalation between Iran and GCC states risks
blockading that strait, triggering a global energy supply shock that would spike
oil prices, destabilize economies worldwide, and set back the renewable energy
transition for years.

This analysis explores GCC energy consumption patterns, fossil fuel dependence,
and what the data tells us about why these nations have every incentive to see
the Iran conflict resolved quickly.

```{r}
#| label: filter-countries

countries <- c("World", "United States", "Europe",
               "Saudi Arabia", "United Arab Emirates", 
               "Qatar", "Kuwait", "Bahrain", "Oman")

gcc <- energy_raw |>
  filter(country %in% countries) |>
  select(
    country, year,
    energy_per_capita,
    fossil_share_energy,
    coal_electricity,
    gas_electricity,
    oil_electricity,
    nuclear_electricity,
    hydro_electricity,
    solar_electricity,
    wind_electricity,
    other_renewable_electricity
  ) |>
  filter(year >= 2000) |>
  mutate(
    region = case_when(
      country %in% c("Saudi Arabia", "United Arab Emirates",
                     "Qatar", "Kuwait", "Bahrain", "Oman") ~ "GCC",
      country == "United States"                           ~ "United States",
      country == "Europe"                                  ~ "Europe",
      country == "World"                                   ~ "World Average"
    )
  )
```
## What the Data Shows

The first chart makes the gap impossible to ignore. Qatar consumes nearly 
200,000 kWh per person per year — roughly **8 times the world average** and 
almost **3 times the United States**. Every GCC country sits well above both 
the US and Europe.

This is not by accident. GCC nations face some of the harshest climate 
conditions on earth, with summer temperatures regularly exceeding 45°C (113°F). 
Air conditioning is not optional — it runs constantly, in homes, offices, 
shopping malls, and even outdoor spaces. On top of that, energy prices in most 
GCC countries are heavily government-subsidized, meaning there is little 
financial pressure on consumers or industries to use less.

The result is a region that is extraordinarily dependent on energy just to 
function — and as the next chart shows, almost all of that energy comes from 
fossil fuels.

```{r}
#| label: chart-per-capita
#| fig-width: 10
#| fig-height: 6

gcc |>
  filter(year == 2022, !is.na(energy_per_capita)) |>
  mutate(country = fct_reorder(country, energy_per_capita)) |>
  ggplot(aes(x = country, y = energy_per_capita, 
             fill = country == "World")) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "grey50", "FALSE" = "#006C35")) +
  scale_y_continuous(labels = comma_format()) +
  coord_flip() +
  labs(
    title   = "Energy Consumption Per Capita: GCC vs World Average (2022)",
    subtitle = "GCC nations consume far more energy per person than the world average",
    x       = NULL,
    y       = "kWh per capita",
    caption = "Source: Our World in Data"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```


```{r}
#| label: chart-fossil-dependence
#| fig-width: 10
#| fig-height: 6

gcc |>
  filter(year == 2022, !is.na(fossil_share_energy)) |>
  mutate(country = fct_reorder(country, fossil_share_energy)) |>
  ggplot(aes(x = country, y = fossil_share_energy, fill = region)) +
  geom_col(show.legend = TRUE) +
  scale_fill_manual(values = c(
    "GCC"           = "#006C35",
    "United States" = "#B22234",
    "Europe"        = "#003399",
    "World Average" = "grey50"
  )) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  coord_flip() +
  labs(
    title    = "Fossil Fuel Dependence by Country (2022)",
    subtitle = "% of total energy consumption coming from coal, oil, and gas",
    x        = NULL,
    y        = "% of energy from fossil fuels",
    fill     = NULL,
    caption  = "Source: Our World in Data"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

Kuwait and Saudi Arabia sit at nearly 100% fossil fuel dependence. Even the 
UAE, which has invested in solar and nuclear, still derives over 90% of its 
energy from fossil fuels. Compare that to Europe at around 70% — still high, 
but on a clear downward trend thanks to renewables and nuclear investment.

This total dependence on fossil fuels creates a dangerous feedback loop: 
burning fossil fuels drives climate change, rising temperatures demand more 
cooling, more cooling requires more energy, and more energy means more 
emissions. The GCC is caught at the center of this cycle — and the Strait of 
Hormuz sits at the center of their ability to sustain it.

```{r}
#| label: wide-to-long

gcc_long <- gcc |>
  filter(year == 2022) |>
  select(country, region,
         coal_electricity, gas_electricity, oil_electricity,
         nuclear_electricity, hydro_electricity, solar_electricity,
         wind_electricity, other_renewable_electricity) |>
  pivot_longer(
    cols      = -c(country, region),
    names_to  = "source",
    values_to = "twh"
  ) |>
  mutate(
    source = source |>
      str_remove("_electricity") |>
      str_replace("other_renewable", "Other Renewables") |>
      str_to_title()
  ) |>
  drop_na(twh)

head(gcc_long)
```
```{r}
#| label: chart-energy-mix
#| fig-width: 10
#| fig-height: 6

source_colors <- c(
  "Coal"               = "#3d3d3d",
  "Oil"                = "#7a5215",
  "Gas"                = "#c06020",
  "Nuclear"            = "#8e44ad",
  "Hydro"              = "#1a6fa8",
  "Solar"              = "#e8b820",
  "Wind"               = "#27ae60",
  "Other Renewables"   = "#5dade2"
)

gcc_long |>
  mutate(country = fct_reorder(country, twh, .fun = sum)) |>
  ggplot(aes(x = country, y = twh, fill = source)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = source_colors) +
  scale_y_continuous(labels = percent_format()) +
  coord_flip() +
  labs(
    title    = "Energy Mix by Source (2022)",
    subtitle = "GCC countries rely almost entirely on gas and oil — renewables are barely visible",
    x        = NULL,
    y        = "Share of electricity generation",
    fill     = "Source",
    caption  = "Source: Our World in Data"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold"),
    legend.position = "bottom"
  )
```
```{r}
#| label: check-columns

energy_raw |>
  filter(country == "Saudi Arabia") |>
  select(year) |>
  bind_cols(
    energy_raw |>
      filter(country == "Saudi Arabia") |>
      select(contains("gdp") | contains("oil") | contains("rent"))
  ) |>
  glimpse()
```
```{r}
#| label: chart-oil-gdp
#| fig-width: 10
#| fig-height: 6

energy_raw |>
  filter(country %in% countries, year == 2021) |>
  mutate(
    region = case_when(
      country %in% c("Saudi Arabia", "United Arab Emirates",
                     "Qatar", "Kuwait", "Bahrain", "Oman") ~ "GCC",
      country == "United States"                           ~ "United States",
      country == "Europe"                                  ~ "Europe",
      country == "World"                                   ~ "World Average"
    )
  ) |>
  filter(!is.na(oil_share_energy)) |>
  mutate(country = fct_reorder(country, oil_share_energy)) |>
  ggplot(aes(x = country, y = oil_share_energy, fill = region)) +
  geom_col(show.legend = TRUE) +
  scale_fill_manual(values = c(
    "GCC"           = "#006C35",
    "United States" = "#B22234",
    "Europe"        = "#003399",
    "World Average" = "grey50"
  )) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  coord_flip() +
  labs(
    title    = "Oil Share of Total Energy Consumption (2021)",
    subtitle = "GCC economies run on oil — a closed Strait cuts off both their exports and their income",
    x        = NULL,
    y        = "% of energy from oil",
    fill     = NULL,
    caption  = "Source: Our World in Data"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold"),
    legend.position = "bottom"
  )
```
```{r}
#| label: chart-chokepoints
#| fig-width: 10
#| fig-height: 6

chokepoints <- tibble(
  location = c(
    "Strait of Hormuz",
    "Strait of Malacca",
    "Suez Canal",
    "Bab el-Mandeb",
    "Turkish Straits",
    "Danish Straits",
    "Cape of Good Hope"
  ),
  barrels_per_day = c(21, 16, 9.2, 8.8, 4.8, 3.4, 1.5),
  involves_gcc = c(TRUE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE)
)

chokepoints |>
  mutate(location = fct_reorder(location, barrels_per_day)) |>
  ggplot(aes(x = location, y = barrels_per_day, 
             fill = involves_gcc)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "#006C35", "FALSE" = "grey60")) +
  geom_text(aes(label = paste0(barrels_per_day, "M bbl/day")),
            hjust = -0.1, size = 3.5, fontface = "bold") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 25)) +
  labs(
    title    = "World's Major Oil Chokepoints by Daily Flow (2023)",
    subtitle = "The Strait of Hormuz carries more oil than any other single point on earth",
    x        = NULL,
    y        = "Million barrels per day",
    caption  = "Source: U.S. Energy Information Administration (EIA), 2023"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

The Strait of Hormuz is not just another shipping lane. It carries more oil 
than any other single point on earth — nearly 21 million barrels every day. 
No other chokepoint comes close. If this strait closes, there is no realistic 
alternative route that can absorb that volume. The world would feel it immediately.

```{r}
#| label: chart-revenue-loss
#| fig-width: 10
#| fig-height: 6

revenue <- tibble(
  country = c("Saudi Arabia", "UAE", "Kuwait", "Qatar", "Oman", "Bahrain"),
  daily_export_revenue_billion = c(5.8, 2.1, 1.5, 1.9, 0.5, 0.1)
)

revenue |>
  mutate(
    country = fct_reorder(country, daily_export_revenue_billion),
    monthly_loss = daily_export_revenue_billion * 30
  ) |>
  ggplot(aes(x = country, y = daily_export_revenue_billion)) +
  geom_col(fill = "#006C35") +
  geom_text(aes(label = paste0("$", daily_export_revenue_billion, "B/day")),
            hjust = -0.1, size = 3.5, fontface = "bold") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 8)) +
  labs(
    title    = "Estimated Daily Oil Export Revenue at Risk (2023)",
    subtitle = "Revenue lost per day if the Strait of Hormuz is closed",
    x        = NULL,
    y        = "Billion USD per day",
    caption  = "Source: EIA, OPEC Annual Statistical Bulletin 2023"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

Saudi Arabia stands to lose nearly $6 billion every single day the Strait is 
closed. Combined across all GCC nations, that is over $12 billion in daily 
revenue gone. These are not economies that can absorb that kind of shock for 
long — oil exports are not one part of their economy, they are virtually the 
entire economy.

```{r}
#| label: chart-import-collapse
#| fig-width: 10
#| fig-height: 6

imports <- tibble(
  country  = c("Saudi Arabia", "UAE", "Kuwait", "Qatar", "Oman", "Bahrain"),
  food_pct        = c(80, 90, 90, 90, 80, 85),
  medicine_pct    = c(75, 85, 80, 85, 70, 80),
  technology_pct  = c(95, 95, 90, 90, 85, 90)
) |>
  pivot_longer(
    cols      = -country,
    names_to  = "category",
    values_to = "pct_imported"
  ) |>
  mutate(
    category = case_when(
      category == "food_pct"       ~ "Food",
      category == "medicine_pct"   ~ "Medicine",
      category == "technology_pct" ~ "Technology"
    )
  )

imports |>
  mutate(country = fct_reorder(country, pct_imported, .fun = mean)) |>
  ggplot(aes(x = country, y = pct_imported, fill = category)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(
    "Food"       = "#c06020",
    "Medicine"   = "#8e44ad",
    "Technology" = "#1a6fa8"
  )) +
  scale_y_continuous(labels = percent_format(scale = 1),
                     limits = c(0, 110)) +
  geom_text(aes(label = paste0(pct_imported, "%")),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 2.8, fontface = "bold") +
  labs(
    title    = "GCC Import Dependency: Food, Medicine & Technology (2023)",
    subtitle = "With oil revenue gone, GCC nations cannot afford to pay for basic necessities",
    x        = NULL,
    y        = "% sourced from imports",
    fill     = NULL,
    caption  = "Source: World Bank, FAO, 2023"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )
```

This is where the vulnerability becomes a survival issue. GCC nations import 
nearly all of their food, medicine, and technology. They can only afford to do 
that because of oil revenue. Without exports flowing through the Strait, that 
purchasing power disappears almost overnight. A country that consumes 8 times 
the world average in energy and imports 90% of its food has no buffer. There 
is no fallback.

## Conclusion

The data tells a clear and connected story. GCC nations consume extraordinary 
amounts of energy — driven by extreme heat, subsidized prices, and energy 
intensive industries like desalination and oil refining. That consumption runs 
almost entirely on fossil fuels, with renewables barely registering in the 
energy mix. And the revenue that sustains these societies flows almost entirely 
through one narrow waterway — the Strait of Hormuz.

A closure of the Strait would not just inconvenience these economies. It would 
cut off the income they depend on to import food, medicine, and basic goods. 
Nations that consume the most energy per person on earth have built societies 
that physically cannot function without it — and cannot pay for it without 
their oil exports moving freely.

This is why the GCC has every reason to want the Iran conflict resolved. It is 
not just about regional stability or diplomacy. It is about the economic and 
physical survival of nations that have built everything around a supply chain 
with a single point of failure.

As global energy demand continues to rise — accelerated by AI data centers, 
EV adoption, and population growth — the pressure on that chokepoint only 
increases. The transition to renewables is not just a climate issue. For the 
GCC, it may eventually be the only path to an economy that does not depend on 
a strait that one country can close overnight.



