---
title:  "Gym Members Exercise Dataset"
author: "Mark Hamer"
date:   today
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
execute:
  echo:    true
  warning: false
---

```{r}
set.seed(42)

library(tidyverse)

project_theme <- theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40"),
    legend.position = "bottom"
  )
```

## Data Source

This dataset was obtained from Kaggle (Vala Khorasani, "Gym Members Exercise
Dataset", 2024: https://www.kaggle.com/datasets/valakhorasani/gym-members-exercise-dataset).
It contains 973 observations of gym members' workout sessions across 15 variables,
including physiological metrics (age, weight, BMI, heart rate), session
characteristics (duration, calories burned, workout type), and self-reported
behaviors (water intake, workout frequency). The dataset was selected from
Discussion 5A and stored as a flat .csv file prior to any transformation.

## Data Structure Before Tidying


```{r ds-load}
gym_raw <- read_csv("gym_members_exercise_tracking.csv")

glimpse(gym_raw)

cat("Rows:", nrow(gym_raw), "| Columns:", ncol(gym_raw), "\n")

head(gym_raw, 6)
```

## Transformation Steps

**Step 1** adds a unique ID to each member so every row can be 
traced back to its original member after reshaping.
```{r}
gym_raw <- gym_raw |>
  mutate(member_id = row_number())
```

**Step 2** renames all 15 columns to clean snake_case, removing 
spaces and parentheses from the original column names.

```{r}
gym_clean <- gym_raw |>
  rename(
    age                = Age,
    gender             = Gender,
    weight_kg          = `Weight (kg)`,
    height_m           = `Height (m)`,
    max_bpm            = Max_BPM,
    avg_bpm            = Avg_BPM,
    resting_bpm        = Resting_BPM,
    session_duration_h = `Session_Duration (hours)`,
    calories_burned    = Calories_Burned,
    workout_type       = Workout_Type,
    fat_pct            = Fat_Percentage,
    water_intake_l     = `Water_Intake (liters)`,
    workout_freq       = `Workout_Frequency (days/week)`,
    experience_level   = Experience_Level,
    bmi                = BMI
  )
```

**Step 3** reshapes from wide to long format using pivot_longer(). 
Each member now has one row per metric instead of one row with 
15 columns.

```{r}
gym_long <- gym_clean |>
  pivot_longer(
    cols      = -c(member_id, gender, workout_type, experience_level),
    names_to  = "metric",
    values_to = "value"
  )
```

**Step 4** normalizes variable types — metric names are lowercased, 
values are forced to numeric, and experience level is converted to 
an ordered factor so plots sort correctly.

```{r}
gym_long <- gym_long |>
  mutate(
    metric           = str_to_lower(metric),
    value            = as.numeric(value),
    experience_level = factor(experience_level,
                              levels = c(1, 2, 3),
                              labels = c("Beginner", "Intermediate", "Advanced"),
                              ordered = TRUE)
  )
```

**Step 5** checks for and removes missing values. The dataset 
contains no NAs — all 11,676 rows are retained after reshaping 
(973 members × 12 numeric metrics each).

```{r}
na_count <- gym_long |>
  summarise(total_missing = sum(is.na(value)))

cat("Missing values before cleaning:", na_count$total_missing, "\n")

gym_tidy <- gym_long |>
  drop_na(value)

cat("Rows after cleaning:", nrow(gym_tidy), "\n")
```


**Analysis prep** pivots back to wide format for plotting and adds 
two derived variables: age_group buckets members into decades, and 
training_load combines session duration and weekly frequency into a 
single weekly volume estimate.

```{r}
gym_wide <- gym_tidy |>
  pivot_wider(names_from = metric, values_from = value)

gym_wide <- gym_wide |>
  mutate(
    age_group = cut(age,
                    breaks = c(18, 30, 40, 50, 100),
                    labels = c("20s", "30s", "40s", "50s+"),
                    right  = FALSE),
    experience_level = factor(experience_level,
                              levels = c(1, 2, 3),
                              labels = c("Beginner", "Intermediate", "Advanced"),
                              ordered = TRUE),
    calories_per_hour = calories_burned / session_duration_h
  )
```

## Average Session Duration by Age Group
```{r}
duration_by_age <- gym_wide |>
  group_by(age_group) |>
  summarise(
    mean_duration = mean(session_duration_h, na.rm = TRUE),
    se            = sd(session_duration_h, na.rm = TRUE) / sqrt(n())
  )

ggplot(duration_by_age,
       aes(x = age_group, y = mean_duration, fill = age_group)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_errorbar(aes(ymin = mean_duration - se,
                    ymax = mean_duration + se),
                width = 0.2) +
  labs(
    title    = "Average Session Duration by Age Group",
    subtitle = "Error bars show ± 1 standard error",
    x        = "Age Group",
    y        = "Mean Session Duration (hours)"
  ) +
  project_theme

```
##Workout Efficiency:

```{r}
ggplot(gym_wide,
       aes(x = workout_type, y = calories_per_hour, fill = workout_type)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c(
    "Cardio"   = "#E63946",
    "HIIT"     = "#2196F3",
    "Strength" = "#FF9800",
    "Yoga"     = "#4CAF50"
  )) +
  labs(
    title    = "Workout Efficiency: Calories Burned per Hour",
    subtitle = "Normalizing for session duration reveals true workout intensity",
    x        = "Workout Type",
    y        = "Calories per Hour"
  ) +
  project_theme
```

**Workout Efficiency:** When calories burned are normalized by session 
duration to produce a calories-per-hour metric, all four workout types 
show nearly identical median efficiency at approximately 720 calories 
per hour. This reveals that the apparent calorie advantage of HIIT and 
Cardio in raw totals is largely explained by longer average session 
durations rather than greater intensity per hour. No single workout 
type is meaningfully more efficient than another in this dataset, 
suggesting that consistency and duration matter more than workout type 
for total calorie expenditure.

**Weekly Training Load:**
```{r}
gym_wide <- gym_wide |>
  mutate(
    age_group = cut(age,
                    breaks = c(18, 30, 40, 50, 100),
                    labels = c("20s", "30s", "40s", "50s+"),
                    right  = FALSE),
    experience_level = factor(experience_level,
                              levels = c(1, 2, 3),
                              labels = c("Beginner", "Intermediate", "Advanced"),
                              ordered = TRUE),
    calories_per_hour = calories_burned / session_duration_h,
    training_load     = session_duration_h * workout_freq
  )
```


```{r}
gym_wide |>
  select(training_load, resting_bpm, fat_pct, bmi) |>
  pivot_longer(
    cols      = -training_load,
    names_to  = "metric",
    values_to = "value"
  ) |>
  mutate(metric = recode(metric,
    "resting_bpm" = "Resting BPM",
    "fat_pct"     = "Fat Percentage",
    "bmi"         = "BMI"
  )) |>
  ggplot(aes(x = training_load, y = value)) +
  geom_point(alpha = 0.3, size = 1.5, color = "#2196F3") +
  geom_smooth(method = "lm", se = TRUE, color = "#E63946", linewidth = 1) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(
    title    = "Weekly Training Load vs Fitness Markers",
    subtitle = "Training load = session duration × weekly frequency",
    x        = "Training Load (hours/week)",
    y        = "Value"
  ) +
  project_theme
```

When combining session duration and workout frequency into a single training 
load score, a clear pattern emerges for body fat. Members with low training 
loads (~2 hours/week) average around 31% body fat, while those with high 
training loads (~10 hours/week) average around 14%. That is a 17 point drop  
the strongest trend in the entire dataset. BMI and resting heart rate show no 
such pattern, suggesting that how much you train per week matters for body 
composition but not for heart rate metrics.


## Conclusions

**Age and Session Duration:** Members of all ages work out for roughly 
the same amount of time — around 1.2 to 1.3 hours per session. Age 
does not affect how long someone stays at the gym.

**Workout Efficiency:** Once you account for session length, all four 
workout types burn about the same calories per hour — around 720. 
HIIT and Cardio only looked better before because those sessions are 
longer on average, not because they work harder. The type of workout 
matters less than how long you do it.

**Weekly Training Load:** Members who train more hours per week have 
significantly lower body fat. Those training around 2 hours per week 
average 31% body fat, while those training around 10 hours per week 
average just 14%. That is a 17 point difference and the clearest 
finding in the dataset. Interestingly, BMI and resting heart rate 
barely change no matter how much someone trains, meaning total weekly 
exercise predicts body fat but not heart rate.